Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-03-09 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
minor edits to intro and conclusion

The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance function including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost of learning the covariance function parameters is $O(mn+m)$ in every step of the optimizer or sampler. This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentiation methods used in modern probabilistic programming frameworks, such as Stan \citep{carpenter2017stan} and others. This is because the memory requirements of automatic differentiation scale with the size of the autodiff expression tree which in direct implementations is simpler for basis function than covariance matrix based approach. The basis function approach also provides an easy way to apply a non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which further improves MCMC efficiency.

While \citet{solin2018hilbert} have fully developed the mathematical theory behind this specific approximation of GPs, further work is needed for its practical implementation in probabilistic programming frameworks. In this paper, the interactions among the key factors of the method such as the number of basis functions, domain of the prediction space, and properties of the true functional relationship between covariates and response variable, are investigated and analyzed in detail in relation to the computational performance and accuracy of the method. Practical recommendations are given for the values of the key factors based on simple diagnostic values and intuitive graphical summaries that encode the recognized relationships. Our recommendations will help users to choose valid and optimized values for these factors, improving computational performance without sacrificing modeling accuracy. The proposed diagnostic can be used to check whether the chosen values for the number of basis functions and the domain of the prediction space are adequate to model the data well.

@@

The remainder of the paper is structured as follows. In Section~\ref{ch4_gp}, we introduce GPs, covariance functions and spectral density functions. In Section~\ref{sec_method}, the reduced-rank approximation to GPs proposed by \citet{solin2018hilbert} is described. In Section~\ref{sec_accuracy}, the accuracy of these approximations under several conditions using analytical and numerical methods is analyzed. Several case studies in which we fit exact and approximate GPs to real and simulated data are provided in Section~\ref{sec_cases}. A brief conclusion of the work is made in Section~\ref{sec_conclusion}. Appendix~\ref{app_approx_covfun} includes a brief presentation of the mathematical details behind the Hilbert space approximation of a stationary covariance function, and Appendix \ref{sec_periodic} presents a low-rank representation of a GP for the particular case of a periodic covariance function. Online supplemental material with more case studies illustrating the performance and applicability of the method can be found online at \url{https://github.com/gabriuma/basis_functions_approach_to_GP} in the subfolder \url{Paper/online_supplemental_material}.

@@

Modeling unknown functions using exact GPs is computationally intractable in many applications. This problem becomes especially severe when performing full Bayesian inference using sampling-based methods. In this paper, a recent approach for a low-rank representation of stationary GPs, originally proposed by \citet{solin2018hilbert}, has been analyzed in detail. The method is based on a basis function approximation via Laplace eigenfunctions. The method has an attractive computational cost as it effectively approximates GPs by linear models, which is also an attractive property in modular probabilistic programming programming frameworks. The dominating cost per log density evaluation (during sampling) is $O(nm+m)$, which is a big benefit in comparison to $O(n^3)$ of a exact GP model. The obtained design matrix is independent of hyperparameters and therefore only needs to be constructed once, at cost $O(nm)$. All dependencies on the kernel and the hyperparameters are through the prior distribution of the regression weights. The parameters' posterior distribution is $m$-dimensional, where $m$ is usually much smaller than the number of observations $n$.

The main contribution of this paper is an in-depth analysis and diagnostic of the performance and accuracy of the approximation in relation to the key factors of the method, that is, the number of basis functions, the boundary condition of the Laplace eigenfunctions, and the non-linearity of the function to be learned. Recommendations for the values of these key factors based on the recognized relations among them have been provided along with illustrations of these relations. These illustrations will not only help users to improve performance and save computation time, but also serve as a powerful diagnostic tool whether the chosen values for the number of basis functions and the boundary condition are adequate to fit to the data at hand with sufficient accuracy.

@@

The obtained functional relationships between the key factors influencing the approximation not only help users to visually assess the accuracy of the method
but can also serve an automatic diagnostic tool.
In this paper, we primarily studied the functional relationships for univariate inputs. Accordingly, investigating the functional relationships more thoroughly for multivariate inputs remains a topic for future research.

>>>>

The Laplace eigenfunctions can be computed analytically and they are independent of the particular choice of the covariance function including the hyperparameters. While the pre-computation cost of the basis functions is $O(m^2n)$, the computational cost of learning the covariance function parameters is $O(mn+m)$ in every step of the optimizer or sampler. This is a big advantage in terms of speed for iterative algorithms such as Markov chain Monte Carlo (MCMC). Another advantage is the reduced memory requirements of automatic differentiation methods used in modern probabilistic programming frameworks, such as Stan \citep{carpenter2017stan} and others. This is because the memory requirements of automatic differentiation scale with the size of the autodiff expression tree which in direct implementations is simpler for basis function than covariance matrix-based approaches. The basis function approach also provides an easy way to apply a non-centered parameterization of GPs, which reduces the posterior dependency between parameters representing the estimated function and the hyperparameters of the covariance function, which further improves MCMC efficiency.

While \citet{solin2018hilbert} have fully developed the mathematical theory behind this specific approximation of GPs, further work is needed for its practical implementation in probabilistic programming frameworks. In this paper, the interactions among the key factors of the method such as the number of basis functions, domain of the prediction space, and properties of the true functional relationship between covariates and response variable, are investigated and analyzed in detail in relation to the computational performance and accuracy of the method. Practical recommendations are given for the values of the key factors based on simple diagnostic values and intuitive graphical summaries that encode the recognized relationships. Our recommendations help users to choose valid and optimized values for these factors, improving computational performance without sacrificing modeling accuracy. The proposed diagnostic can be used to check whether the chosen values for the number of basis functions and the domain of the prediction space are adequate to model the data well. On that basis, we also develop an iterative procedure to achieve accuracte approximation performance with minimal computational costs.

@@

The remainder of the paper is structured as follows. In Section~\ref{ch4_gp}, we introduce GPs, covariance functions and spectral density functions. In Section~\ref{sec_method}, the reduced-rank approximation to GPs proposed by \citet{solin2018hilbert} is described. In Section~\ref{sec_accuracy}, the accuracy of these approximations under several conditions is studied using both analytical and numerical methods.  Practical diagnostics are developed there as well. Several case studies in which we fit exact and approximate GPs to real and simulated data are provided in Section~\ref{sec_cases}. A brief conclusion of the work is made in Section~\ref{sec_conclusion}. Appendix~\ref{app_approx_covfun} includes a brief presentation of the mathematical details behind the Hilbert space approximation of a stationary covariance function, and Appendix \ref{sec_periodic} presents a low-rank representation of a GP for the particular case of a periodic covariance function. Online supplemental material with more case studies illustrating the performance and applicability of the method can be found online at \url{https://github.com/gabriuma/basis_functions_approach_to_GP} in the subfolder \url{Paper/online_supplemental_material}.

@@

Modeling unknown functions using exact GPs is computationally intractable in many applications. This problem becomes especially severe when performing full Bayesian inference using sampling-based methods. In this paper, a recent approach for a low-rank representation of stationary GPs, originally proposed by \citet{solin2018hilbert}, has been analyzed in detail. The method is based on a basis function approximation via Laplace eigenfunctions. The method has an attractive computational cost as it effectively approximates GPs by linear models, which is also an attractive property in modular probabilistic programming programming frameworks. The dominating cost per log density evaluation (during sampling) is $O(nm+m)$, which is a big benefit in comparison to $O(n^3)$ of an exact GP model. The obtained design matrix is independent of hyperparameters and therefore only needs to be constructed once, at cost $O(nm)$. All dependencies on the kernel and the hyperparameters are through the prior distribution of the regression weights. The parameters' posterior distribution is $m$-dimensional, where $m$ is usually much smaller than the number of observations $n$.

As one of the main contributions of this paper, we provided an in-depth analysis of the approximation's performance and accuracy in relation to the key factors of the method, that is, the number of basis functions, the boundary condition of the Laplace eigenfunctions, and the non-linearity of the function to be learned. On that basis, as our second main contribution, we developed practical diagnostics to assess the approximation's performance as well as an iterative procedure to obtain an accurate approximation with minimal computational costs.

%We illustrated our findings in several case studies of real and simulated data. 
%%Recommendations for the values of these key factors based on the recognized relations among them have been provided. These illustrations will not only help users to improve performance and save computation time, but also serve as a powerful diagnostic tool whether the chosen values for the number of basis functions and the boundary condition are adequate to fit to the data at hand with sufficient accuracy.

@@

In this paper, the obtained functional relationships between the key factors influencing the approximation and corresponding diagnostics were studied primarily for univariate inputs. Accordingly, investigating the functional relationships more thoroughly for multivariate inputs remains a topic for future research.
